Persistence in functional topology: local-connectivity, q-tameness and a correction to a theorem of Morse

### 2021-04-09 connectivity.tex
Add historical stuff from Max thesis

For more results of this type, we refer to \cite{MR1481706}. There, the above proposition \cite[Corollary VI.12.6]{MR1481706} and many similar ones are proven using sheaf theoretic methods.
For example, Bredon also discusses cohomology local connectedness \cite[Section II.17]{MR1481706} and various other comparison results, e.g.\@ between singular and Borel-Moore homology \cite[Corollary V.12.15]{MR1481706} in the presence of homology local connectedness, as well as examples such as spaces that are sheaf cohomology locally connected but not homology locally connected \cite[Example II.17.12]{MR1481706}.

@@

We will now show that the weak notion does not imply q-tameness in general.
To do so let us consider the \textit{$d$-dimensional Hawaiian earring}
\begin{equation*}
\HE = \bigcup_{n\in\mathbb{N}}\left\{(x_0,\dots,x_d)\in\R^{d+1} \ \middle | \ \left(x_0-\frac{1}{n}\right)^2+x_1^2+\dots+x_d^2=\left(\frac{1}{n}\right)^2\right\}.
\end{equation*}

@@

\begin{thm} \label{thm:counterexample}
	Let $\H$ be a homology theory such that its value on a point is finitely generated for every degree and on $\HE$ is not finitely generated for some degree. Then, the function $f \colon \HE \to \R$ whose value at the origin is $0$ and is $1$ everywhere else defines a $\HLC$ Morse filtration
	\todo{weakly?}
	that is not q-tame.
\end{thm}

@@

\begin{proof}
	The case of singular homology is well know and can be found in \cite{Barratt.1962}.
	For \v{C}ech homology we use the fact that it commutes with inverse limits for compact Hausdorff spaces.
	Define 
	\begin{align*}
	\HE_k &= \left\{(x_0,\dots,x_d)\in\R^{d+1}\mid \left(x_0-\frac{1}{k}\right)^2+x_1^2+\dots+x_d^2=\left(\frac{1}{k}\right)^2\right\}\\
	&\cup\bigcup_{n=1}^{k-1}\left\{(x_0,\dots,x_d)\in\R^{d+1}\mid \left(x_0-\frac{1}{n}\right)^2+x_1^2+\dots+x_d^2=\left(\frac{1}{n}\right)^2\right\},
	\end{align*}
	i.e., the $d$-dimensional Hawaiian earring but with the $k$-th largest $d$-sphere filled.
	\todo{inequality?}
	We have $\lim_{k}\HE_{k} = \bigcap_{k}\mathbb{H}^{d}_{k}=\mathbb{H}^{d}$, and hence $\CH_{d}(\HE) = \lim_{k}\CH_{d}(\mathbb{H}^{d}_{k})$.
	One can easily check that each $\mathbb{H}^{d}_{k}$ satisfies the assumptions for \cref{prop:cech_sing_hom_hlc}, which implies $\lim_{k}\CH_{d}(\mathbb{H}^{d}_{k})=\lim_{k}H_{d}(\mathbb{H}^{d}_{k})$.
	We compute
	\begin{equation*}
	\lim_{k}H_{d}(\mathbb{H}^{d}_{k})=\lim\left(\dots\to \prod_{n=1}^2\mathbb{F}\to \prod_{n=1}^1\mathbb{F}\to \prod_{n=1}^0\mathbb{F}\right)=\prod_{n\in\mathbb{N}}\mathbb{F},
	\end{equation*}
	which is infinite-dimensional over $\mathbb{F}$.
	This finishes the proof.
\end{proof}

@@

	We will now show that the class $\Sigma_{s,t}$ is closed under finite unions.
	For $i \in \{1, 2\}$ let $K_i$ be in $\Sigma_{s,t}$ with $\delta_i > 0$ and $K_i \subseteq U_i$ open such that $U_{i} \cap X_{s+\delta_i} \to L \cap X_{t}$ is $\HS_n$.
	We Use Lemma \ref{l:neighborhood third} to construct sets $K_i^\prime$ such that
	\begin{equation*}
	K_i \subseteq \interior(K_i^\prime) \subseteq K_i^\prime \subseteq U_i.
	\end{equation*}
	Notice that for $\delta = \min(\delta_i)$ we have $K_i^\prime \cap X_{s+\delta} \to L \cap X_t$ is $\HS_n$.
	Additionally, the induction hypothesis implies that $K_1 \cap K_2 \cap X_s \to K_1^\prime \cap K_2^\prime \cap X_{s+\delta}$ is $\HS_{(n-1)}$.
	We therefore have the following commutative diagram satisfying the assumptions of Lemma~\ref{l:commutative algebra}:
	\begin{equation*}
	\begin{tikzcd}
	\H_n(L \cap X_t) \oplus \H_n(L \cap X_t) \arrow[r] &
	\H_n(L \cap X_t) & \\
	\H_{n}(K_1^\prime \cap X_{s+\delta}) \oplus \H_n(K_2^\prime \cap X_{s+\delta}) \arrow[r] \arrow[u] & 
	\H_{n}((K_1^\prime \cap X_{s+\delta}) \cup (K_2^\prime \cap X_{s+\delta})) \arrow[r] \arrow[u] &
	\H_{n-1}(K_1^\prime \cap K_2^\prime \cap X_{s+\delta}) \\ & 
	\H_{n}((K_1 \cup K_2) \cap X_s) \arrow[r] \arrow[u] &
	\H_{n-1}(K_1 \cap K_2 \cap X_s). \arrow[u]
	\end{tikzcd}
	\end{equation*}
	We conclude that $K_1 \cup K_2 \in \Sigma_{s, t}$.
	Since any compact $K \subseteq \interior(L)$ can be expressed as a finite union of sets in $\Sigma_{s,t}$ the induction step and the lemma are proven.
\end{proof}

>>>>

For more results of this type, we refer to \cite{MR1481706}. There, the above proposition \cite[Corollary VI.12.6]{MR1481706} and many similar ones are proven using sheaf theoretic methods.
For example, Bredon also discusses cohomology local connectedness \cite[Section II.17]{MR1481706} and various other comparison results, e.g.\@ between singular and Borel--Moore homology \cite[Corollary V.12.15]{MR1481706} in the presence of homology local connectedness, as well as examples such as spaces that are sheaf cohomology locally connected but not homology locally connected \cite[Example II.17.12]{MR1481706}.

@@

We will now show that the weak notion does not imply q-tameness in general.
To do so let us consider the \textit{$d$-dimensional Hawaiian earring}
\begin{equation*}
\HE = \bigcup_{n\in\mathbb{N}}\left\{(x_0,\dots,x_d)\in\R^{d+1} \ \middle | \ \left(x_0-\frac{1}{n}\right)^2+x_1^2+\dots+x_d^2=\left(\frac{1}{n}\right)^2\right\}.
\end{equation*}
\todo{credit to Milnor}

@@

\begin{thm} \label{thm:counterexample}
	Let $\H$ be a (generalized) homology theory such that its value on a point is finitely generated for every degree and on $\HE$ is not finitely generated for some degree. Then, the function $f \colon \HE \to \R$ whose value at the origin is $0$ and is $1$ everywhere else defines a weakly $\HLC$ Morse filtration that is not q-tame.
\end{thm}

@@

\begin{proof}
	The case of singular homology is well know and can be found in \cite{Barratt.1962}.
	For \v{C}ech homology we use the fact that it commutes with inverse limits for compact Hausdorff spaces.
	Define 
	\begin{align*}
	\HE_k &= \left\{(x_0,\dots,x_d)\in\R^{d+1}\mid \left(x_0-\frac{1}{k}\right)^2+x_1^2+\dots+x_d^2 \leq \left(\frac{1}{k}\right)^2\right\}\\
	&\cup\bigcup_{n=1}^{k-1}\left\{(x_0,\dots,x_d)\in\R^{d+1}\mid \left(x_0-\frac{1}{n}\right)^2+x_1^2+\dots+x_d^2=\left(\frac{1}{n}\right)^2\right\},
	\end{align*}
	i.e., the $d$-dimensional Hawaiian earring but with the $k$-th largest $d$-sphere filled.
	We have $\lim_{k}\HE_{k} = \bigcap_{k}\mathbb{H}^{d}_{k}=\mathbb{H}^{d}$, and hence $\CH_{d}(\HE) = \lim_{k}\CH_{d}(\mathbb{H}^{d}_{k})$.
	Clearly, each $\mathbb{H}^{d}_{k}$ is a CW-complex, so we can simply use cellular homology to compute
	\begin{equation*}
	\lim_{k}\CH_{d}(\mathbb{H}^{d}_{k})=\lim\left(\dots\to \prod_{n=1}^2\mathbb{F}\to \prod_{n=1}^1\mathbb{F}\to \prod_{n=1}^0\mathbb{F}\right)=\prod_{n\in\mathbb{N}}\mathbb{F},
	\end{equation*}
	which is infinite-dimensional over $\mathbb{F}$.
	This finishes the proof.
\end{proof}

@@

	We will now show that the class $\Sigma_{s,t}$ is closed under finite unions.
	For $i \in \{1, 2\}$ let $K_i$ be in $\Sigma_{s,t}$ with $\delta_i > 0$ and $K_i \subseteq U_i$ open such that $U_{i} \cap X_{s+\delta_i} \to L \cap X_{t}$ is $\HS_n$.
	We use Lemma \ref{l:neighborhood third} to construct sets $K_i^\prime$ such that
	\begin{equation*}
	K_i \subseteq \interior(K_i^\prime) \subseteq K_i^\prime \subseteq U_i.
	\end{equation*}
	Notice that for $\delta = \min(\delta_i)$ we have $K_i^\prime \cap X_{s+\delta} \to L \cap X_t$ is $\HS_n$.
	Additionally, the induction hypothesis implies that $K_1 \cap K_2 \cap X_s \to K_1^\prime \cap K_2^\prime \cap X_{s+\delta}$ is $\HS_{(n-1)}$.
	We therefore have the following commutative diagram satisfying the assumptions of Lemma~\ref{l:commutative algebra}:\todo{check assumptions for mayer-vietoris to apply}
	\begin{equation*}
	\begin{tikzcd}
	\H_n(L \cap X_t) \oplus \H_n(L \cap X_t) \arrow[r] &
	\H_n(L \cap X_t) & \\
	\H_{n}(K_1^\prime \cap X_{s+\delta}) \oplus \H_n(K_2^\prime \cap X_{s+\delta}) \arrow[r] \arrow[u] & 
	\H_{n}((K_1^\prime \cap X_{s+\delta}) \cup (K_2^\prime \cap X_{s+\delta})) \arrow[r] \arrow[u] &
	\H_{n-1}(K_1^\prime \cap K_2^\prime \cap X_{s+\delta}) \\ & 
	\H_{n}((K_1 \cup K_2) \cap X_s) \arrow[r] \arrow[u] &
	\H_{n-1}(K_1 \cap K_2 \cap X_s). \arrow[u]
	\end{tikzcd}
	\end{equation*}
	We conclude that $K_1 \cup K_2 \in \Sigma_{s, t}$.
	Since any compact $K \subseteq \interior(L)$ can be covered by a finite union of sets in $\Sigma_{s,t}$ the induction step and the lemma are proven.
\end{proof}

### 2021-04-09 historical.tex
Add historical stuff from Max thesis

>>>>

\todo{Max: the following subsections are taken verbatim from my thesis and not adapted to this paper yet} 
\subsection{Critical Points}
So far, we have mostly talked about changes in the homology of sublevel sets at certain \emph{values} since this is what is studied in persistent homology. The original motivation to study Functional Topology was, however, the study of critical \emph{points}, which we briefly touched upon in section \ref{sec:morse}.

In \cite{Morse.1939}, Morse and Tompkins make the following definition.

\begin{defi}
Let $(M,d)$ be a metric space and $f\colon M\to\mathbb{R}$ a function. We say $p\in M$ is a \emph{homotopically ordinary point of $f$} if there is a neighborhood $U$ of $p$ in $f_{\leq f(p)}$ and a continuous map $H\colon U\times[0,1]\to M$ such that
\begin{itemize}
\item $H(q,0)=q$ for all $q$,
\item $H(p,1)\neq p$ and
\item for any $e>0$ there exists $\delta>0$ with the property that whenever $s<s'$ and $d(H(q,s),H(q,s'))>e$ hold, we have $f(H(q,s))-f(H(q,s'))>\delta$.
\end{itemize}
We say that $p$ is a \emph{homotopically critical point of $f$} if it is not homotopically ordinary.
\end{defi}

Roughly, one can think of homotopically critical points of $f$ as points $p$ that have no neighborhood in $f_{\leq f(p)}$ that can be mapped by a homotopy into $f_{\leq t}$ for some $t<f(p)$.

In the case of isolated critical points, one can then choose a neighborhood $U$ of a homotopically critical point $p$ not containing any other critical point and define the \emph{cap type numbers of $p$} as the cap type numbers of the relative homology lattice associated to the restriction of $f$ to this neighborhood. The total cap type numbers are then defined as the sums of all of these cap type numbers. Assuming a finite number of critical points, Morse and Tompkins then proceed to state Morse inequalities for these total cap type numbers, referring to \cite{Morse.1940}, which was still in preparation at that time, for proof.

More generally, in the non-isolated case, they consider closed sets of critical points possessing a neighborhood that does not contain any other critical points and call these \emph{critical sets}. Again, one can then define the cap type numbers of such sets as the cap type numbers of $f$ restricted to the corresponding neighborhood. In order to get Morse inequalities in this setting, without assuming that the critical points are finite in number, Morse and Tompkins additionally assume local-$f$-connectedness and again refer to \cite{Morse.1940} for proof. However, the proofs that Morse presents there repeatedly make use of his erroneous claim that local-$f$-connectedness implies q-tameness, so they cannot be seen as valid.

\subsection{The Douglas Functional}
After the previously described general statements on critical points, Morse and Tompkins then consider the following setting introduced in even greater generality by Douglas: Let $g=(g_i)\colon\mathbb{R}\to\mathbb{R}^n$ be a simple closed curve such that $g$ is differentiable and such that each $g'_i$ is Lipschitz. Assume that $g$ is $2\pi$-periodic. Let $\tilde{\Omega}$ be the space of continuous functions $\varphi\colon\mathbb{R}\to\mathbb{R}$ with $\varphi((t)+2\pi)=\varphi(t)+2\pi$ for all $t$ and such that there are three distinct points $\alpha_i\in[0,2\pi)$ with $\varphi(\alpha_i)=\alpha_i$. We define the \emph{Douglas-functional} associated to the curve $g$ on $\tilde{\Omega}$ by $$A_g(\varphi)=\frac{1}{16}\int_0^{2\pi}\int_0^{2\pi}\sin\left(\frac{\alpha-\beta}{2}\right)^{-2}\lVert g(\varphi(\alpha))-g(\varphi(\beta)) \rVert_2^2\mathrm{d}\alpha\mathrm{d}\beta$$ and set $$\Omega_g=\{\varphi\in\tilde\Omega\mid A_g(\varphi)<\infty\}.$$ Douglas proved that, for even more general $g$ than above, $\Omega_g$ is non-empty and that sublevel sets of $A_g$ on $\Omega_g$ are compact. Since $A_g$ is clearly bounded below by $0$, this implies that $A_g$ has a unique global minimizer. Douglas then used harmonic analysis to show that this minimizer of $A_g$ corresponds to a solution of Plateau's Problem. More generally, each critical point of this functional $A_g$ corresponds to a critical point of the area functional.

Now, Morse and Tompkins first assume that $A_g$ has finitely many homotopically critical points. They show that any homotopically critical point of $A_g$ is also a critical point in the classical sense. They verify the other assumptions they make for the Morse inequalities and then conclude via these inequalities that if $A_g$ has two distinct minimizers, it must also have a critical point of index 1, i.e. with non-zero first cap type number. This type of result is often called a Mountain Pass Lemma.

In order to remove the finiteness assumptions, they go on to prove local-$A_g$-connectedness for $\Omega_g$ and require that the two minimizers are contained in disjoint critical sets. Again using the Morse inequalities, they infer the existence of an unstable minimal surface. As we have mentioned before, Morse's proof in \cite{Morse.1940} of the Morse inequalities in this general case is faulty since local-$A_g$-connectedness is not enough to imply the necessary q-tameness. However, the proof of local-$A_g$-connectedness in \cite{Morse.1939} actually shows that $\Omega_g$ is strongly locally-$A_g$-connected. In fact, Morse and Tompkins even show that each sublevel set of $A_g$ is locally contractible in all larger sublevel sets. So if our previous conjectures are true, the general theory can be applied and their approach still works.

### 2021-04-09 homology.tex
Consolidate sections connectivity, counterexample and theorem
Add historical stuff from Max thesis

\subsection{Homology theories}

\begin{itemize}
	\item Definition
	\item Examples, main: singular with integer coeff and cech with field 
	\item Mayer-Vietoris
\end{itemize}

>>>>

\section{Homology theories}

\begin{itemize}
\item Definition (generalized homology theory, without dimension axiom, but with Milnor's additivity axiom)
\item Examples, main: singular with integer coeff and cech with field \cite{Kelly.1961}
\item mention Dowker duality
\item Mayer-Vietoris
\end{itemize}

\todo{Uli: I saw a discussion about the topological invariance of Vietoris homology for non-compact metric spaces.
I thought that it suggested that invariance holds only for compact spaces,
but I can't find it anymore.

What I found was this:
\url{http://pldml.icm.edu.pl/pldml/element/bwmeta1.element.zamlynska-8f369e96-b1eb-4ec9-99e2-8c366367a5a4} claims invariance also for non-compact.
\url{https://www.ams.org/journals/tran/1947-062-02/S0002-9947-1947-0024128-6/S0002-9947-1947-0024128-6.pdf} mentions a non-invariant property "null-bounding"
}

### 2021-04-09 persistence.tex
Add historical stuff from Max thesis

\begin{itemize}
	\item Persistence module
	\item p.f.d
	\item q-tameness
	\item Barcode
	\item bottleneck
	\item stability
\end{itemize}

>>>>

\todo{start with the standard vector space setting, before generalizing?}
Let $R$ be a ring. The category of \emph{persistence $R$-modules} is the functor category  $(\Mod{R})^{\mathbf{R}}$, where $\mathbf{R}$ is the category corresponding to the real numbers $\mathbb{R}$ considered as a poset. We will simply say \emph{persistence module} if the coefficient ring is clear from the context.

A persistence module $M$ is called \emph{pointwise finitely generated (p.f.g.)} if $M_{t}$ is a finitely generated module for every $t\in\mathbb{R}$. $M$ is called \emph{q-tame} if $\im (M_{s}\to M_{t})$ is finitely generated for all $s<t\in\mathbb{R}$.

The theory of persistence modules is particularly nice if $R=\mathbb{F}$ is a field, which we will assume for the rest of this section. In this case, we will say that $M$ is \emph{pointwise finite-dimensional (p.f.d.)} instead of p.f.g.

What makes the field coefficient case nice is that it allows for a combinatorial description of many persistence modules in terms of basic summands. Concretely, we say that $M$ has a \emph{barcode} if there is a family of intervals $(I_{\alpha})_{\alpha}$, $I_{\alpha}\subseteq\mathbb{R}$ such that
\[
M\cong\bigoplus_{\alpha} C(I_{\alpha}),
\]
where the persistence module $C(I)$ for an interval $I$ is defined via
\[
C(I)_t=
\begin{cases}
    \mathbb{F} & \text{if } t\in I,\\
    0              & \text{otherwise,}
\end{cases}
\qquad
\text{with structure maps}
\qquad
C(I)_{t,u}=
\begin{cases}
    \operatorname{id}_{\mathbb{F}} & \text{if } t,u\in I,\\
    0              & \text{otherwise.}
\end{cases}
\]

Interval modules are indecomposable in the category of persistence modules and by the Krull--Remak--Schmidt--Azumaya Theorem \cite{MR37832}, two direct sums of interval modules can only be isomorphic if the corresponding families of intervals already agree. Hence, we can speak of \emph{the} barcode of a persistence module and use it as a complete isomorphism invariant.

Unfortunately, not every persistence module has a barcode, so general existence results are important to have. The most widely used of these is Crawley-Boevey's Theorem \cite{Crawley-Boevey.2015}, which asserts that every p.f.d.\ persistence module has a barcode.

As is well-known, Crawley-Boevey's Theorem does not extend to q-tame persistence modules. However, there are still ways to get a notion of barcode for q-tame persistence modules, for example by phrasing the definition of barcode in terms of so-called rectangle measures \cite{Chazal.2016a}. Other methods include using a slightly weaker notion of isomorphism for the direct sum decomposition into interval modules \cite{Chazal.2016b} or imposing additional algebraic constraints and allowing for the decompositions to be in terms of products instead of sums \cite{schmahl2020structure}. For each of these alternative methods, restricting to the p.f.d.\ setting recovers the basic notion of barcode discussed before.

The usefulness of barcodes in applications often rests on their stability. To make precise what stability means, we define the \emph{interleaving distance} between two persistence modules $M$ and $N$ as 
\[
d_{I}(M,N)=\inf\{\delta>0\mid\text{there exists a }\delta\text{-interleaving between }M\text{ and }N\},
\]
where a $\delta$-interleaving between $M$ and $N$ is given by a collection of maps $M_{t}\to N_{t+\delta}$ and $N_{t}\to M_{t+\delta}$ for each $t\in\mathbb{R}$ such that the diagrams
\[
\begin{tikzcd}
M_s\arrow[r]\arrow[d] & M_t\arrow[d] \\
N_{s+\delta}\arrow[r] & N_{t+\delta}
\end{tikzcd}
\qquad
\begin{tikzcd}
N_s\arrow[r]\arrow[d] & N_t\arrow[d] \\
M_{s+\delta}\arrow[r] & M_{t+\delta}
\end{tikzcd}
\qquad
\begin{tikzcd}
M_{t-\delta}\arrow[rd]\arrow[r]& M_{t}\arrow[rd]\arrow[r] & M_{t+\delta} \\
N_{t-\delta}\arrow[ru]\arrow[r]& N_{t}\arrow[ru]\arrow[r] & N_{t+\delta}
\end{tikzcd}
\]
commute for all $s<t\in\mathbb{R}$.

We also define the \emph{Bottleneck distance} between two barcodes $B=(I_{\alpha})_{\alpha\in A}$ and $B'=(I'_{\alpha'})_{\alpha'\in A'})$ as
\[
d_{b}(B,B')=\inf\{\delta>0\mid\text{there exists a }\delta\text{-matching between }M\text{ and }N\},
\]
where a $\delta$-matching between $B$ and $B'$ is given by subsets $X\subseteq A$, $X'\subseteq A'$ and a bijection $\sigma\colon X\to X'$ such that
\[
\lvert \sup I - \inf I \rvert < 2\delta
\]
whenever there is $\alpha\in A\setminus X$ with $I=I_{\alpha}$ or $\alpha'\in A'\setminus X'$ with $I=I'_{\alpha'}$, and
\begin{align*}
\lvert \sup I'_{\sigma(\alpha)} - \sup I_{\alpha} \rvert &< \delta \\
\lvert \inf I'_{\sigma(\alpha)} - \inf I_{\alpha} \rvert &< \delta
\end{align*}
for all $\alpha\in X$.

\begin{thm}[\cite{}]\todo{what do we want to cite here?}
Let $M$ and $M'$ be q-tame persistence modules with barcodes $B$ and $B'$. Then
\[
d_{b}(B,B')= d_{I}(M,M').
\]
\end{thm}

The most commonly studied example of a persistence module is \emph{persistent homology}, i.e., the persistence module obtained by applying homology to an $\mathbb{R}$-indexed diagram of spaces. In what follows, we will study the persistent homology of certain classes of such diagrams, namely filtrations.
